---
title: "Smart Beta - Fundamental Indexing"
author: "Rohan Thanki and Anirudh Bhat"
date: "10/19/2021"
output: pdf_document
classoption: a4paper

header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 10,
  fig.height = 6,
  tidy.opts = list(width.cutoff = 100),
  tidy = TRUE,
  message = FALSE,
  warning = FALSE
  )
```

# Project Aim

* The objective of this assignment is to construct an equity indices as an alternative to a traditional market capitalization weighted index and compare the performance of these created indices. 
* These indices are created based on fundamental variables from the financial statements of these firms (available in COMPUSTAT data) and ratios computed from these fundamental variables

# Initial Setup

## Loading libraries

```{r}

# libraries to read and write date to a file
library(data.table)
library(readxl)
library(writexl)
library(xlsx)

# libraries to fetch data using APIs
library(fredr)

#libraries to wrangle data and perform analytics
library(tidyverse)
library(zoo)
library(lubridate)
library(stringr)
library(ppsr)
library(moments)
library(sqldf)
library(DescTools)
library(naniar)
library(PerformanceAnalytics)
library(tidyquant)

# libraries to plot data
library(corrplot)
```

## Declaring Global Variables

```{r}
# Data path
data_path = "../Data/"

# add this to a config file instead
fredr_set_key("7a257e710452eb46c904661109420fe8")

crsp_mkt_cap_threshold = 1000000 # actually this represents 1000 million since shares outstanding are in thousands
crsp_price_threshold <- 5

# this uses the hard-drive as an extended memory if the RAM is full
invisible(utils::memory.limit(64000))
```

* Due to memory constraints, we will only load data of CUSIPS that are common to both CRSP and COMPUSTAT data
* I have created this list in SAS and saved it as a CSV file

# Data Preparation

## Unique CUSIPS for consideration

```{r}
# compustat <- fread('data/compustat_funda_filtered.csv')
# compustat_cusips <- compustat %>%
#   rename(CUSIP = cusip) %>%
#   mutate(CUSIP = substr(CUSIP, 1, 8)) %>% 
#   select(CUSIP) %>% 
#   unique()
# 
# crsp_msf <- fread('data/crsp_msf_filtered.csv')
# crsp_msf_cusips <- crsp_msf %>%
#   filter(SHRCD %in% c(10,11)) %>% 
#   select(CUSIP) %>% 
#   unique()
# 
# common_cusips <- compustat_cusips %>% 
#   inner_join(crsp_msf_cusips, by = 'CUSIP')
# 
# write.csv(common_cusips, 'data/common_cusips.csv', row.names = FALSE)
```

```{r}
# Selecting MSFT, AAPL, FORD
# cusip_ids <- c('59491810', '03783310', '34537086')
cusip_ids <- fread("../Data/common_cusips.csv")$CUSIP
```

## COMPUSTAT - Annual Data - Company Fundamentals

* link - `https://www.marketplace.spglobal.com/en/datasets/compustat-fundamentals-(8)#:~:text=Compustat%20Fundamentals%20provides%20standardized%20North,on%20for%20over%2050%20years`

Compustat Fundamentals provides standardized North American and global financial statement and market data for over 80,000 active and inactive publicly traded companies as far back as 1950 and point-in-time snapshots beginning in 1987. The dataset includes:

* over 3,000 standardized annual, quarterly, year-to-date, and semi-annual financial items covering income statement, balance sheet, cash flow, segment (for North American companies), and ratio data
* Industry specific data covering 12 industries*
* Daily and monthly market data, including security pricing, dividends, and splits
* Corporate Tracker providing corporate action and security identifier history
* Point-in-time snapshot data for North American companies providing all changes to fundamental items and their associated point dates

### Dealing with Look-Ahead Bias

* Here, we need to deal with "look-ahead bias", i.e. we should not form our investment decisions based on information we do not have at that moment of time.
* Since the COMPUSTAT data has companies whose fiscal years end at different months in the same year, we get the `Year` value from the `fyear` column. We then lag this data by a year to form portfolios for the next calendar year
* For example - all the financials of companies released in `fiscal year 2001` are used to form portfolios for `calendar year 2002`

```{r}
compustat <- fread(paste0(data_path, "compustat_funda_filtered.csv"), showProgress = FALSE)

compustat_w <- compustat %>% 
  select(c('cusip', 'fyear', 'at', 'lct', 'dlc', 'dltt', 'teq', 'oancf', 'ivncf', 'fincf', 'scf', 'ibc', 'xidoc', 'dpc', 'txdc', 'esubc', 'sppiv', 'fopo', 'fsrco', 'exre', 'sale', 'dv', 'capx', 'ivch', 'aqc', 'fuseo', 'sppe', 'siv', 'ivstch', 'ivaco', 'oiadp', 're', 'prcc_f', 'csho', 'lt', 'act', 'ni', 'pi', 'dp')) %>%
  arrange(cusip, fyear) %>% 
  mutate(cusip = substr(cusip, 1, 8)) %>%
  filter(cusip %in% cusip_ids) %>% 
  filter(!is.null(cusip)) %>%
  rename(Year = fyear, CUSIP = cusip) %>% 
  select(CUSIP, Year, everything()) %>%
  distinct(CUSIP, Year, .keep_all = TRUE) %>%
  group_by(CUSIP) %>%
  fill(everything(), .direction = "downup") %>% 
  mutate(
    
    # Lagging the whole time series by 1 year to prevent look-ahead bias
    Year = lead(Year),
    
    # Computing Variables
    'Net_Assets' = rollapply(at - lct, 5, mean, align='right', fill=NA),
    'Book_Capitalization' = rollapply(dlc + dltt + teq, 5, mean, align='right', fill=NA),
    'CFO' = rollapply(oancf, 5, mean, align='right', fill=NA),
    'CF_Total' = rollapply(oancf + ivncf + fincf, 5, mean, align='right', fill=NA),
    'Internal_Cash_Flow' = rollapply(case_when(scf %in% c(1,2,3) ~ ibc+xidoc+dpc+txdc+esubc+sppiv+fopo+fsrco,
                                          scf == 7 ~ ibc+xidoc+dpc+txdc+esubc+sppiv+fopo+exre),
                                     5, mean, align='right', fill=NA),
    'Revenue' = rollapply(sale, 5, mean, align='right', fill=NA),
    'Dividend' = rollapply(dv, 5, mean, align='right', fill=NA),
    'Investment' = rollapply(case_when(scf %in% c(1,2,3) ~ capx+ivch+aqc+fuseo-sppe-siv,
                         scf == 7 ~ capx+ivch+aqc-sppe-siv-ivstch-ivaco),
                         5, mean, align='right', fill=NA),
    'Profitability' = rollapply(oiadp/((at + lag(at))/2), 5, mean, align='right', fill=NA),
    'Asset_Turnover' = rollapply(sale/at, 5, mean, align='right', fill=NA),
    
    # Compute more fundamental variables

    # Altman Z-score
    AZ_A = (act-lct)/at,
    AZ_B = re/at,
    AZ_C = oiadp/at,
    AZ_D = (prcc_f*csho)/lt,
    AZ_E = sale/at,
    Altman_Z_score = 1.2*AZ_A + 1.4*AZ_B + 3.3*AZ_C + 0.6*AZ_D + 0.99*AZ_E,
    
    # Ohlson O-score
    OHL_A = log(at),
    OHL_B = lt/at,
    OHL_C = (act - lct)/at,
    OHL_D = lct/act,
    OHL_E = ni/at,
    OHL_F = (pi+dp)/lt,
    OHL_G = case_when( lt > at ~ 1, TRUE ~ 0),
    OHL_H = case_when(ni < 0 & lag(ni) < 0 ~ 1, TRUE ~ 0),
    OHL_I = ((ni - lag(ni))/(abs(ni) + abs(lag(ni)))),
    Ohlson_O_score = -1.32 -(0.407*OHL_A) + (6.03*OHL_B) - (1.43*OHL_C) + (0.0757*OHL_D) - (2.37*OHL_E) - (1.83 * OHL_F) - (1.72*OHL_G) + (0.285*OHL_H) - (0.521*OHL_I)
    ) %>% 
  
  # removing intermediate columns
  select(-at:-dp, -AZ_A:-AZ_E, -OHL_A:-OHL_I)

rm(compustat)
```

## CRSP Data

* The Center for Research in Security Prices (CRSP) database contains end-of-day and month-end price on primary listings for the NYSE, NYSE MKT, NASDAQ, and Arca exchanges along with basic market indices. 
* The CRSP_MSF data contains all end-of-month information on stocks from the CRSP over the period 1926 - Dec 2019

### Dealing with Look-Ahead Bias

* Here, we need to deal with "look-ahead bias", i.e. we should not form our investment decisions based on information we do not have at that moment of time.
* Since we mostly use the CRSP data to `compute returns`, we do not lag the data.
* However, we do lag the `Market Cap` data by 1 month, since this is used to compute ratios that affect our investment decisions. For example, the `average market cap from December 2000 to November 2001` are used are used to form portfolios for `calendar year 2002`
* This need not be the case for practical purposes, since market caps can be obtained in real time for most publicly traded stocks. However, since CRSP releases its data with a 1 month lag, we lag the market cap by 1 month in this case.

### Loading and Wrangling MSF Data

```{r}
# Loading CRSP Monthly Data
crsp_msf <- fread(paste0(data_path, "crsp_msf_filtered.csv"), showProgress = FALSE) %>%
  select(CUSIP, DATE, PRC, RET, SHROUT, VWRETD, SHRCD) %>%
  filter(CUSIP %in% cusip_ids, SHRCD %in% c(10,11))

# Account for Bankrupt Companies
crsp_msf[crsp_msf$RET == 'B']$RET <- -1

# Wrangling CRSP Monthly Data
crsp_msf_w <- crsp_msf %>% 
  rename(Date = DATE) %>%
  select(Date, everything()) %>%
  group_by(CUSIP) %>% 
  fill(everything(), .direction = "downup") %>%  
  mutate(Date = as.yearmon(ymd(Date)),
         Year = year(Date),
         RET = as.numeric(RET),
         VWRETD = as.numeric(VWRETD),
         PRC = abs(PRC),
         MKT_CAP = lag(PRC) * lag(SHROUT),
    
    ) %>%
  distinct(CUSIP, Date, .keep_all = TRUE) %>%
  drop_na(RET) %>% 
  select(-PRC,-SHROUT, -SHRCD) %>%
  select(CUSIP, Date, Year, everything()) %>% 
  arrange(CUSIP,Date)

rm(crsp_msf)
```

## Combining COMPUSTAT and CRSP Data

```{r}
crsp_compustat_combined <- crsp_msf_w %>%
  inner_join(compustat_w, by = c('CUSIP', 'Year')) %>%
  select(CUSIP, Year, everything())
```

## Fama French Factor Returns

* The Fama/French factors are constructed using the 6 value-weight portfolios formed on size and book-to-market.
* SMB (Small Minus Big) is the average return on the three small portfolios minus the average return on the three big portfolios
* HML (High Minus Low) is the average return on the two value portfolios minus the average return on the two growth portfolios
* Rm-Rf, the excess return on the market, value-weight return of all CRSP firms incorporated in the US and listed on the NYSE, AMEX, or NASDAQ that have a CRSP share code of 10 or 11 at the beginning of month t, good shares and price data at the beginning of t, and good return data for t minus the one-month Treasury bill rate (from Ibbotson Associates)

```{r}
# Fama French Returns Data
ff_3_factor <- fread("../Data/F-F_Research_Data_Factors.csv", skip = 3) %>%
  rename(Date = V1,
         MKT = 'Mkt-RF') %>%
  mutate(Date = as.yearmon(as.character(Date), "%Y%m"),
         MKT = MKT / 100,
         SMB = SMB / 100,
         HML = HML / 100,
         RF = RF / 100)
```

## FRED Data

We extract the following data from the `Federal Reserve Economic Data (FRED)` database. The data we will require is:

1. NBER Recession Data
  *# This time series is an interpretation of US Business Cycle Expansions and Contractions data provided by The National Bureau of Economic Research (NBER). 
  * NBER identifies months and quarters of turning points without designating a date within the period that turning points occurred. The dummy variable adopts an arbitrary convention that the turning point occurred at a specific date within the period. 
  * A value of 1 is a recessionary period, while a value of 0 is an expansionary period.
  
2. Kansas City Financial Stress Index (KCFSI)
  * The Kansas City Financial Stress Index (KCFSI) is a monthly measure of stress in the U.S. financial system based on 11 financial market variables.
  * A positive value indicates that financial stress is above the long-run average, while a negative value signifies that financial stress is below the long-run average. Another useful way to assess the current level of financial stress is to compare the index to its value during past, widely recognized episodes of financial stress.

```{r}
# NBER Recession Data
nber <- fredr(series_id = "USREC") %>%
  rename(Date = date,
         USREC = value) %>%
  mutate(Date = as.yearmon(Date)) %>%
  select(Date, USREC)

# KCFSI Data
kcfsi <- fredr(series_id = "KCFSI") %>%
  rename(Date = date,
         KCFSI = value) %>%
  mutate(Date = as.yearmon(Date)) %>%
  select(Date, KCFSI)

# Merging Data
fred_data <- nber %>%
  full_join(kcfsi, by = 'Date')
```

# Computing Weights of Stock for Each Portfolio

## Check if Stock Exists for Subsequent Year

* Since stocks can get listed anytime in the year, we can only consider them in our portfolios for the subsequent year
* The `existence_check` dataframe stores the values of the years in which each stock is eligible to be included in our portofolios

```{r}
existence_check <- crsp_msf_w %>% 
  mutate(Month = month(Date)) %>% 
  filter(Month == 12) %>% 
  select(CUSIP, Year) %>%
  arrange(CUSIP, Year) %>% 
  group_by(CUSIP) %>%
  slice(2:n())
```

## Computing Yearly Weights Based on Each Fundamental Variable

* While computing weights, we ignore stocks with negative values of ratios - so their weights are set to 0

```{r}
col_sum_divide <- function(x)
  return(x/sum(x))

compute_weights <- function(df) {
  weights <- df %>% 
    mutate(across(c(-CUSIP), col_sum_divide))
  return(weights)
}

weights_df <- crsp_compustat_combined %>% 
  inner_join(existence_check, by=c('CUSIP', 'Year'))

weights_df[weights_df < 0] = 0
weights_df <- weights_df %>% 
  mutate_all(function(x) ifelse(!is.finite(x), 0, x)) 

weights_df <- weights_df %>% 
  select(Year, CUSIP, Net_Assets:Asset_Turnover) %>% 
  group_by(Year) %>%
  group_modify(~compute_weights(.))

# For debugging
# fwrite(weights_df, 'output/weights.csv')
```
 
# Computing Returns for Each Portfolio

## Monthly Returns

### Weighted Return for each Stock for each Month

```{r}
# Joining the Weights dataframe with the Returns Dataframe
weighted_returns_df <- weights_df %>% 
  inner_join(crsp_msf_w[, c('CUSIP', 'Date', 'Year', 'RET', 'VWRETD')], by=c('CUSIP', 'Year')) %>% 
  ungroup() %>% 
  select(-Year) %>% 
  select(Date, CUSIP, RET, VWRETD, everything()) %>% 
  mutate(across(Net_Assets:Asset_Turnover, ~ .x * RET))

# For debugging
# fwrite(weighted_returns_df, 'output/weighted_returns.csv')
```


### Monthly Returns for Each Portfolio

```{r}
# Summing Up the Weighted Returns for each date
port_ret_df <- weighted_returns_df %>% 
  group_by(Date) %>% 
  summarize(across(c('Net_Assets':'Asset_Turnover'),
                   .f = list(sum = ~ sum(., na.rm = TRUE))))
```

### Combining Returns 

```{r}
vwretd <- crsp_msf_w %>%
  ungroup() %>% 
  select(Date, VWRETD) %>% 
  distinct()

returns_df_combined <- port_ret_df %>% 
   inner_join(vwretd, by=c('Date'))

# Modifying ugly column names
colnames(returns_df_combined)<-gsub("_sum","",colnames(returns_df_combined))

# Writing the returns to a file
# fwrite(returns_df_combined, 'output/Monthly_Returns.csv')
```

# Combining Data for Plotting

## Monthly Returns

```{r}
plot_df <- returns_df_combined %>% 
  left_join(ff_3_factor, by = 'Date') %>% 
  left_join(fred_data, by = 'Date')

fwrite(plot_df, 'output/Plot_Data.csv')
```

## Annualising Returns
 
```{r}
returns_df_combined_annualized <- plot_df %>% 
  select(Date, Net_Assets:RF) %>% 
  mutate(Year = year(Date)) %>% 
  select(-Date) %>%
  group_by(Year) %>% 
  summarize(across(everything(),
                   .f = list(annualized = ~ Return.annualized(., scale = 12, geometric = TRUE))))

# Modifying ugly column names
colnames(returns_df_combined_annualized)<-gsub("_annualized","",colnames(returns_df_combined_annualized))

fwrite(returns_df_combined_annualized, './output/Annualized_Return.csv')
```

## Cumulative Returns for Each Portfolio from an arbitary Start Year

```{r}
start_year = 2009
end_year = 2020
port_cum_ret_df <- returns_df_combined %>%
  filter(year(Date) >= start_year) %>%
  mutate(across(c(-Date), .f = list(cum = function(x) {cumprod(1+x)} )))

# Modifying ugly column names
port_cum_ret_df <- port_cum_ret_df %>% 
  select(Date, contains('cum'))
colnames(port_cum_ret_df)<-gsub("_cum","",colnames(port_cum_ret_df))

# For debugging
# fwrite(port_cum_ret_df, 'output/Cumulative_Returns.csv')
```

# Portfolio Performance Evaluation

```{r}
start_year = 1975
end_year = 2020
performance_metrics <- returns_df_combined %>%
  inner_join(ff_3_factor[, c('Date', 'MKT', 'SMB', 'HML', 'RF')], by = 'Date') %>% 
  filter(year(Date) >= start_year, year(Date) <= end_year) %>% 
  pivot_longer(cols = c(Net_Assets:VWRETD),
               names_to = 'Portfolio',
               values_to = 'RET') %>% 
  select(-Date) %>% 
  group_by(Portfolio) %>% 
  summarize(across(c(RET),
                   .f = list(volatility = ~ sd(., na.rm = T) * sqrt(12),
                             skewness = ~ Skew(., na.rm = T) / (12 ** 1.5),
                             kurtosis = ~ (Kurt(., na.rm = T) / (12 ** 2)) + 3,
                             Sharpe_Ratio = ~ mean(. - RF, na.rm=T) / (sd(., na.rm = T) * sqrt(12)),
                             Information_Ratio = ~ mean(. - MKT, na.rm=T) / (sd(. - MKT, na.rm = T) * sqrt(12)) )))

colnames(performance_metrics)<-gsub("RET_","",colnames(performance_metrics))

fwrite(performance_metrics, './output/Portfolio_Metrics.csv')
```




















